---
title: "Exercise: Anti-religious speech"
author: "DL Oberski and L Boeschoten"
format: 
  html:
    theme: zephyr
    toc: true
    code-tools: true
    code-link: true
editor: visual
execute:
  cache: false
---

## Exercises

```{r, warning=FALSE, message=FALSE}

library(tidyverse)
library(broom)
library(poLCA)
```

Read the data from the General Social Survey 1987. It's not old, it's a classic!

```{r}
antireli <- read.csv("https://daob.nl/files/lca/antireli_data.csv")

head(antireli)
```

Show the data as pattern frequencies.

```{r}
table(antireli) |> knitr::kable()
```

**Question 1**

Use `poLCA` and fit a two-class LCA to these data.

```{r}
fit <- poLCA(cbind(Y1, Y2, Y3) ~ 1, 
             data = antireli, 
             nclass = 2)
```

**Question 2**

Create a profile plot.

```{r}
plot(fit)
```

In this case the default plot is still somewhat readable, but in practice it is not the best as data visualizations go. A simple line plot does a better job (in my personal & completely subjective opinion!) and allows you to display confidence intervals to boot. We use `tidy` from the `broom` package to extract the results and `ggplot` to plot.

```{r}
tidy(fit) %>% 
  filter(outcome == 2) %>% 
  mutate(class = as.factor(class)) %>%
  ggplot(aes(variable, estimate, group = class, color = class)) +
  geom_point() + geom_line() + 
  geom_errorbar(aes(ymin = estimate - 2*std.error, 
                    ymax = estimate + 2*std.error), width = 0.2) +
  theme_bw() +  scale_color_brewer(palette = "Set2")
```

**Question 3**

How would you label the classes?

**Question 4**

What can you say about the estimated class sizes? What does this mean for the prevalence of the attitudes you labeled under (3)?

**Question 5**

Model fit

-   a\. How many parameters are there?

-   b\. How many *unique* data patterns are there (fixing the sample size $n=1713$)?

-   c\. Can you explain the number of degrees of freedom?

-   d\. Can you explain the value of the `G^2` ($G^2$) and `X^2` ($\chi^2$) statistics?
