---
title: "Exercise: Attitudes towards climate change in Europe"
author: "DL Oberski & L Boeschoten"
format: 
  html:
    theme: zephyr
    toc: true
    code-tools: true
    code-link: true
editor: visual
execute:
  cache: true
---

## Exercise: Attitudes towards climate change in Europe

You can use code snippets from the earlier labs to help answer the questions of this exercise. In addition, there is some example code on the bottom of this document that could help you as well.

**Question 1**

Read in the ESS round 10 (2020) climate change attitudes data, you can find the data [here](https://lauraboeschoten.github.io/LCA_GESIS/Extra/Day_2/ess10_climate.csv.gz).

An easy to read codebook copied from ESS is [here](https://lauraboeschoten.github.io/LCA_GESIS/Extra/Day_2/ESS10_codebook.html). The full documentation is [here](https://ess-search.nsd.no/en/study/172ac431-2a06-41df-9dab-c1fd8f3877e7).

-   `ccnthum` - Climate change caused by natural processes, human activity, or both
-   `ccrdprs` - To what extent feel personal responsibility to reduce climate change
-   `wrclmch` - How worried about climate change
-   `testic37` - Imagine large numbers of people limit energy use, how likely reduce climate change
-   `testic38` - How likely, large numbers of people limit energy use
-   `testic39` - How likely, governments in enough countries take action to reduce climate change
-   `gndr` - Gender
-   `agea` - Age of respondent, calculated
-   `eisced` - Highest level of education, ES - [ISCED](https://ec.europa.eu/eurostat/statistics-explained/index.php?title=International_Standard_Classification_of_Education_(ISCED))

*Note*: The data have been preprocessed by ruthlessly subjecting them to `na.omit`. I have also recoded `eisced` to be missing except for values 1-7. Otherwise, the data are as-is from the ESS website.

**Question 2**

In order not to spend most of your precious time waiting, filter the data to only include one country of your choice.

**Question 3**

Perform any exploratory data analyses you find necessary.

**Question 4**

Using `poLCA`, fit LCA models in which the seven participation items are used as indicators (so, exclude `agea`, `gndr`, and `eisced` from the analysis for now). Try models with a different number of classes. *Advice*: try 1--6.

**Question 5**

Use appropriate global fit measures, or any other criteria you prefer, to select the number of classes. Explain your choice.

**Question 6**

Look at local fit measures to assess the fit of your selected model.

**Question 7**

Create a profile plot for your selected model. (*Hint*: You can use the adjusted plotting code below.)

**Question 8**

Interpret your selected model by looking at the profiles. How would you label the classes?

**Question 9**

Create a classification table.

**Question 10**

Calculate the classification error and entropy $R^2$.

**Question 11**

Refit your selected model, now while predicting class membership from `agea`, the square of `agea`, `gndr`, and `eisced`.

**Question 12**

Plot the probability of each class as a function of `agea`, `gndr`, and `eisced`, according to your model. What do you conclude?

**Question 13**

*BONUS*: Investigate the distribution of classes over countries by redoing the analyses using all countries in the ess dataset

**Question 14**

*BONUS*: Deal more appropriately with missing data, for example by using `mice`. You will need the original data from ESS.

### Code hints

Useful libraries

```{r, message=FALSE, warning=FALSE}
set.seed(202303)

library(tidyverse)
library(broom) 
library(haven)
library(poLCA)
```

Read the data from the European Social Survey, round 10 (2020).

```{r, message=FALSE}
ess10_climate <- read_csv("https://daob.nl/files/lca/ess10_climate.csv.gz") 

ess10_climate |> rmarkdown::paged_table()
```

Code to create a subset, in this case of Italy, and fit a model.

```{r}
ess10_climate_it <- filter(ess10_climate, cntry == "IT")
ess10_climate_it$ccrdprs <- ess10_climate_it$ccrdprs + 1

fit <- poLCA(cbind(ccnthum, ccrdprs , wrclmch , 
                     testic37, testic38, testic39) ~ 
                 agea + I(agea^2) + gndr + eisced, 
                data = ess10_climate_it, nclass = 3, 
                maxiter = 2e3, nrep=10, verbose = FALSE)
```

Code to create the profile plot (note that the assignment here differs from the lab)

```{r}
tidy(fit) %>% # from `broom` package
    mutate(class = as.factor(class), outcome = as.factor(outcome)) %>%
    ggplot(aes(outcome, estimate, group = class, color = class)) +
    geom_point() + geom_line() + facet_wrap(~variable, scales = "free_x")+
    geom_errorbar(aes(ymin = estimate - 2*std.error, 
                      ymax = estimate + 2*std.error), width = 0.2) +
    theme_bw() + scale_color_brewer(palette = "Set2")
```

### Effects help

Unfortunately, `effects` does not appear to function properly for this type of model. The code below could be helpful to create effects plots by hand. It assumes that the right-hand side of formula used was `agea + I(agea^2) + gndr + eisced`.

The code below creates a dataframe with the "effects" of the various covariates based on the model estimates from `fit`. This is also how `effects` works and demonstrated within the `poLCA` help file.

```{r}
# Extract posterior probabilities of each class
posterior_df <- as.data.frame(fit$posterior)

# Add covariates to the dataset
posterior_df <- ess10_climate_it %>%
  dplyr::select(agea) %>%
  bind_cols(posterior_df)

# Rename class probability columns
colnames(posterior_df)[2:ncol(posterior_df)] <- paste0("Class_", 1:3)

# Reshape data to long format for ggplot
posterior_long <- posterior_df %>%
  pivot_longer(cols = starts_with("Class_"), 
               names_to = "Class", values_to = "Probability")

# Plot probability of class membership as a function of age
ggplot(posterior_long, aes(x = agea, y = Probability, color = Class)) +
  geom_smooth(method = "loess", se = TRUE) +
  labs(title = "Effect of Age on Class Membership",
       x = "Age", y = "Probability of Class Membership") +
  theme_minimal()
```
